Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-02-15 cbea_revised.tex
Final manuscript

However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}. The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}. Even though these problems are challenging, a very approachable solution is to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden. As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses. Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.

There exists a diverse set of methods already developed in this field. Traditional methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold when trying to select genes for the candidate list. Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change. Of these approaches, methods such as GSEA \cite{subramanian2005} performs a test for each gene set at the population level, summarizing information across all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation. In addition to being able to screen for enriched sets per sample, this strategy also allows for the flexible incorporation of different downstream analyses, such as fitting prediction models, or performing dimension reduction.

@@

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum. However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Some studies have provided emprical performance vealuations supporting this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.

@@

In the following sections, we present our approach titled competitive balances for taxonomic erichment analysis (CBEA). First, we present the step-by-step formulation of CBEA and discuss its statistical properties. Second, we detail our evaluation strategy using both real data and parametric simulations, and the methods we're comparing against. Third, we present results on enrichment testing using CBEA for single samples as well as at the population level. Fourth, we show the performance of CBEA in downstream disease prediction. Finally, we discuss our results and the limitations of our method. An R package implementation of CBEA can be found on GitHub (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/CBEA}).

@@

    \item $\mathbf{X}$: $n$ by $p$ matrix of positive propotions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation. Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix. 

@@

\subsection*{Properties of CBEA}
\subsubsection*{CBEA and balances of groups of parts}
The CBEA statistic is based on the multiplication-based aggregation approach used to calculate balances between groups of parts \cite{egozcue2005}. These balances are computed using the isometric log ratio (ILR) transformation \cite{egozcue2003} formula. For a given balance $i$ splitting variables across sets $R$ and $S$, we have the balance coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$.

CBEA belongs to a set of methods that seek to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}. Unlike methods such as PhILR \cite{silverman2017}, CBEA does not provide a sequential binary partition that forms the basis for ILR procedure \cite{egozcue2003} and is therefore not a subclass of ILR. A similar method to CBEA would be phylofactor \cite{washburne2017}. However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances apriori using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa. This formulation aligns with the $Q_1$ null hypothesis from the gene set testing literature \cite{tian2005}.

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, as suggested by Efron \cite{efron2004}.

@@

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a two-component normal mixture distribution. Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

@@

\subsubsection*{Variance inflation due to inter-taxa correlation}  
When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated. Without loss of generalizability, for a set of taxa with taxa-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:  
\begin{equation} \label{vareq:1}
    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
\end{equation}
where $\sigma_i$ is the standard deviation of taxon $i$ and $\rho_{ij}$ is the correlation between $i$ and $j$. The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation. The CBEA statistic follows a similar pattern. Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write CBEA score for a set $k$ with size $K$ as follows:  
\begin{equation}\label{vareq:2}
    M_{i,k} = \sqrt{\frac{K(p - K)}{K + (p - K)}} \left( \overbar{\log{X_{i,j|j \in K}}} - \overbar{\log{X_{i,j|j \notin K}}} \right)   
\end{equation}
where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$. The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions. As such, the pooled variance of CBEA is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$. The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.

@@

\subsubsection*{Sample-level inference} 
Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set. We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations. Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa. For this evaluation, we estimated type I error as the fraction of samples where our random set is detected as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.  Additionally, this analysis also tests whether CBEA is sensitive to different set sizes.

@@

\item 16S rRNA gene sequencing of stool samples from IBD patients in the pediatric RISK cohort \cite{gevers2014}. This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada. Of the 671 samples obtained, 500 samples belong to patients with IBD. We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU that was present (at least 1 count) in 10\% of the samples or less.   
\end{enumerate}

@@

We also evaluated phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using assigned continuous scores (instead of performing a hypothesis test). This aspect is captured through computing AUROC values comparing computed enrichment scores and true labels. Consistent with prior approaches, adjusting for correlation did not improve performance, where the AUROC values are equivalent to using the Wilcoxon Rank sum statistic at around 0.5. Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.

@@

\subsection*{Disease Prediction}   
Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we want to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.

@@

We fit our model to two data sets with a similar disease classification task of discriminating patients who are diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition. The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al. \cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.

@@

\subsection*{Inference with CBEA} 
CBEA is a microbiome-specific approach to generate sample specific enrichment scores for taxonomic sets defined apriori. The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}. Since this null hypothesis is self-contained per sample, this allows users perform enrichment testing at the sample level. Additionally, in combination with a difference in means test, CBEA can also test for enrichment at the population level across case/control status similar to GSVA \cite{hanzelmann2013}.

For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating respectable power (Fig~\ref{fig:4}). This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution. Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power. Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power. 
For the population-level inference task, CBEA also performed very well. Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error at the correct $\alpha$ threshold while also keeping respectable power. Since the population level enrichment test is equivalent to a differential abundance test using set-based features, we compared the CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}. Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation. We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}). Results for CBEA approaches were replicated in simulation analyses, however for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).

We hypothesized that the discrepancy between simulation and real data evaluations can can arise due to differences in our assumptions regarding the data generating process that informed our simulation schema. For the non-zero component of each taxa, we sampled from the same negative binomial distribution where designated enriched taxa were generated with inflated means (but the same dispersion). These marginals were simulated to account for block exchangable correlation within the enriched set only. This might affect our results in two ways. First, our simulation scenario ensures that all designated non-enriched taxa is identical to each other. This is not the case for real data, where our null scenario involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa. This is represented in \nameref{S7_Fig}, where the distribution of type I error across 500 replications are right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution. Second, as described in the \nameref{introduction} section, we did not consider taxon-specific biases that distort the observed relative abundance of taxa compared to true values \cite{mclaren2019}. In the context of sum-based aggregations, the resulting bias of the aggregated taxon are dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}). Conceptually, this means that measurement error for a taxon aggregate is different across samples as relative abundance of contributing changes, leading to issues when attempting to perform inference. As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our multiplication based approach. This also explains why conversely in simulation studies, where taxon-specific biases are absent, corncob and DESeq2 performed better.

@@

We fit a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs. Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA. Additionally, we also evaluated CBEA with the approach where counts of a set were aggregated using sums and applied the centered log-ratio transformation (CLR). This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}. Results showed that CBEA generate scores perform well across both real data settings and simulation scenarios. Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that CBEA scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses. Comparatively, CBEA generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets. This is consistent with our sample ranking analysis (Fig~\ref{fig:4}), where CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts. However, CBEA did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data. Fortunately, in simulation studies, this performance gap between CLR and CBEA decreases with higher sparsity and correlation, especially in low effect saturation scenarios.

@@

These above results demonstrate the applicability of CBEA under different data analysis scenarios. If researchers are interested in perform inference, they can decide between an unsupervised sample level approach (i.e. screen samples for enrichment of certain characteristics) or a supervised population level approach (i.e. identifying characteristics that are differentially abundant across case/control status). For the unsupervised approach, utilizing the unadjusted CBEA with the mixture normal distribution provides a good initial starting point. In the case where researchers only want to screen samples with mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), they can apply the adjusted approach, which can be effective at conserving type I error even for high correlation scenarios. However, the trade off for this adjustment is power, which decreases with increasing correlation. For the supervised analysis, all CBEA variants control for type I error and provide adequate statistical power. However, using raw CBEA scores with difference-in-means test such as Welch's t-test is preferable since is the least computationally expensive (no estimation process) while still outperforming the use of a sum-based approach with a standard differential abundance test.

>>>>

However, there are unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}. The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}. Even though these problems are challenging, a very approachable solution is to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden. As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses. Furthermore, through the usage of functionally informative sets defined \emph{a priori} based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.

A diverse set of methods have already been developed in this field. Traditional methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold when trying to select genes for the candidate list. Aggregate score methods, which are generally preferred \cite{irizarry2009}, instead assign a score for each gene set based on gene-specific statistics such as z-scores or fold change. Of these approaches, methods such as GSEA \cite{subramanian2005} perform a test for each gene set at the population level, summarizing information across all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation. In addition to being able to screen for enriched sets per sample, this strategy also allows for the flexible incorporation of different downstream analyses, such as fitting prediction models, or performing dimension reduction.

@@

In microbiome research, even when no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum. However, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Some studies have provided emprical performance vealuations supporting this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxon can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.

@@

In the following sections, we present our approach titled competitive balances for taxonomic erichment analysis (CBEA). First, we present the step-by-step formulation of CBEA and discuss its statistical properties. Second, we detail our evaluation strategy using both real data and parametric simulations, and the methods we are comparing against. Third, we present results on enrichment testing using CBEA for single samples as well as at the population level. Fourth, we show the performance of CBEA in downstream disease prediction. Finally, we discuss our results and the limitations of our method. An R package implementation of CBEA can be found on GitHub (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/CBEA}).

@@

    \item $\mathbf{X}$: $n$ by $p$ matrix of positive proportions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation. Users can generate a dense matrix $X$ using a method of choice, however by default mode CBEA will add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix. 

@@

\subsection*{Properties of CBEA}
\subsubsection*{CBEA and balances between groups of parts}
The CBEA statistic is based on the multiplication-based aggregation approach used to calculate balances between groups of parts \cite{egozcue2005}. These balances are computed using the isometric log ratio (ILR) transformation \cite{egozcue2003} formula. For a given balance $i$ splitting variables across sets $R$ and $S$, we have the balance coordinate $x^{*}_{i}$ as: 
\begin{equation}\label{ilr_standard}
    x^{*}_i = \sqrt{\frac{rs}{r+s}} \log\left(\frac{g(X_{j|j \in R})}{g(X_{j|j \in S})}\right)
\end{equation}
where $r$ and $s$ are the cardinalities of sets $R$ and $S$ respectively, $g(z)$ is the geometric mean, and $X_{j}$ are values of the original predictors with indexes defined by membership in $R$ and $S$.

CBEA belongs to a set of methods that seek to leverage compositional balances for the analysis of microbiome data \cite{washburne2017, rivera-pinto2018, silverman2017, morton2017}. Unlike methods such as PhILR \cite{silverman2017}, CBEA does not present an orthonormal basis for the complete ILR transformation (such as a a sequential binary partition) \cite{egozcue2003}. Therefore, it is not a subclass of the ILR transformation and is adjacent to this approach. A similar method to CBEA would be phylofactor \cite{washburne2017}. However, instead of performing an optimization procedure to identify interesting balances, CBEA constructs balances \emph{a priori} using pre-defined sets, and formulates the enrichment of a set as the scaled log-ratio between the center of the subcomposition represented by microbes within the set and the center of the subcomposition represented by remainder taxa. This formulation aligns with the $Q_1$ null hypothesis from the gene set testing literature \cite{tian2005}.

@@

This assumption is also supported by preliminary simulation studies (detailed below). We simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness in Fig~\ref{fig:1}A. We found that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, as suggested by Efron \cite{efron2004}.

@@

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a two-component normal mixture distribution. The goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null is shown in Fig~\ref{fig:1}B. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

@@

\subsubsection*{Variance inflation due to inter-taxa correlation}  
When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated. Without loss of generalizability, for a set of taxa with taxon-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:  
\begin{equation} \label{vareq:1}
    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
\end{equation}
where $\sigma_i$ is the standard deviation of taxon $i$ and $\rho_{ij}$ is the correlation between $i$ and $j$. The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation. The CBEA statistic follows a similar pattern. Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write CBEA score for a set $k$ with size $K$ as follows:  
\begin{equation}\label{vareq:2}
    M_{i,k} = \sqrt{\frac{K(p - K)}{K + (p - K)}} \left( \overbar{\log{X_{i,j|j \in K}}} - \overbar{\log{X_{i,j|j \notin K}}} \right)   
\end{equation}
where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$. The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions. As such, the pooled variance of CBEA is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$. The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.

@@

\subsubsection*{Sample-level inference} 
Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set. We evaluated this application by generating one random taxon-set of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations. Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample, sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa. For this evaluation, we estimated type I error as the fraction of samples where our random set is detected as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.  Additionally, this analysis also tests whether CBEA is sensitive to different set sizes.

@@

\item 16S rRNA gene sequencing of stool samples from IBD patients in the RISK cohort \cite{gevers2014}. This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada. Of the 671 samples obtained, 500 samples belong to patients with IBD. We processed this data set by removing all samples with less than 1,000 total read counts as well as any OTU that was present (at least 1 count) in 10\% of the samples or less.   
\end{enumerate}

@@

We also evaluated phenotype relevance by assessing whether enriched sets according to ground truth are preferentially ranked higher using assigned continuous scores (instead of performing a hypothesis test). This aspect is captured through computing AUROC values comparing computed enrichment scores and true labels. Consistent with the previous type I error evaluation, adjusting for correlation did not improve performance, where obtained AUROC were around 0.5 and at the same level as the benchmark Wilcoxon rank sum statistic. Unadjusted methods were much better at ranking true enriched sets, however the mean AUROC values are lower than alternate single sample enrichment methods (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}) even though this difference is not significant due to overlapping confidence intervals.

@@

\subsection*{Disease Prediction}   
Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we wanted to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.

@@

We fit our model to two data sets with a similar disease classification task of discriminating patients who were diagnosed with IBD (includes both Crohn's disease and ulcerative colitis) using only microbiome taxonomic composition. The two data sets represent different microbiome sequencing aprpaoches: the Gevers et al. \cite{gevers2014} data set uses 16S rRNA gene sequencing, while the Nielsen et al \cite{nielsen2014} data set uses whole genome shotgun sequencing.

@@

\subsection*{Inference with CBEA} 
CBEA is a microbiome-specific approach to generate sample specific enrichment scores for taxonomic sets defined \emph{a priori}. The formulation of CBEA as a comparison between taxa within the set and its complement corresponds to the competitive null hypothesis in the gene set testing literature \cite{tian2005}. Since this null hypothesis is self-contained per sample, this allows users perform enrichment testing at the sample level. Additionally, in combination with a difference in means test, CBEA can also test for enrichment at the population level across case/control status similar to GSVA \cite{hanzelmann2013}.

For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating adequate power (Fig~\ref{fig:4}). This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution. Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power. Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power. 
For the population-level inference task, CBEA also performed very well. Under the permutation global null, representing genera abundance using CBEA scores in combination with Welch's t-test controls for type I error at the correct $\alpha$ threshold while also keeping respectable power. Since the population level enrichment test is equivalent to a differential abundance test using set-based features, we compared the CBEA approach against using element-wise summations with corncob \cite{martin2020} and DESeq2 \cite{love2014} to test for set-level differential abundance. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}. Alternately, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation. We observed that using this approach resulted in an inflated type I error compared to all variants of CBEA (Fig~\ref{fig:2}), yet did not improve power (Fig~\ref{fig:4}). Results for CBEA approaches were replicated in simulation analyses, however for corncob and DESeq2 we observed an opposite effect: in simulation experiments, both methods show good type I error control but low power (\nameref{S2_Fig}, \nameref{S4_Fig}).

We hypothesized that the discrepancy between simulation and real data evaluations could be due to differences in our assumptions regarding the data generating process that informed our simulation schema. For the non-zero component of each taxon, we sampled from the same negative binomial distribution where designated enriched taxa were generated with inflated means (but the same dispersion). These marginals were simulated to account for block exchangable correlation within the enriched set only. This might have affected our results in two ways. First, our simulation scenario ensures that all designated non-enriched taxa are identical to each other. This is not the case for real data, where our null scenario involves randomly sampled sets that might by chance all have taxa with inflated means compared to remainder taxa. This is represented in \nameref{S7_Fig}, where the distribution of type I error across 500 replications is right skewed for underperforming CBEA variants, indicating that these approaches are much more sensitive compared to the Wilcoxon rank sum test or unadjusted CBEA with mixture normal distribution. Second, as described in the \nameref{introduction} section, we did not consider taxon-specific biases that distort the observed relative abundance of taxa compared to true values \cite{mclaren2019}. In the context of sum-based aggregations, the resulting bias of the aggregated taxon-set is dependent on the relative abundances of the contributing taxa (Appendix I in \cite{mclaren2019}). Conceptually, this means that measurement error for a taxon-set is different across samples as relative abundance of contributing taxa changes, leading to issues when attempting to perform inference. As such, we expect methods like corncob or DESeq2 when performed on such sum aggregates in the presence of taxon-specific biases to have inflated type I error compared to our multiplication based approach. This also explains why conversely in simulation studies, where taxon-specific biases are absent, corncob and DESeq2 performed better.

@@

We fit a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using CBEA generated scores as inputs. Similar to our inference analysis, we compared CBEA against both ssGSEA and GSVA. Additionally, we also evaluated CBEA with the approach where counts of a set were aggregated using sums and applied the centered log-ratio transformation (CLR). This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}. Results showed that CBEA generate scores perform well across both real data and simulation scenarios. Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that CBEA scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses. Comparatively, CBEA generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets. This is consistent with our sample ranking analysis (Fig~\ref{fig:4}), where CBEA scores are on average more informative when used to rank samples based on their propensity to have inflated counts. However, CBEA did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data. Fortunately, in simulation studies, this performance gap between CLR and CBEA decreases with higher sparsity and correlation, especially in low effect saturation scenarios.

@@

These above results demonstrate the applicability of CBEA under different data analysis scenarios. If researchers are interested in performing inference, they can decide between an unsupervised sample level approach (i.e. screen samples for enrichment of certain characteristics) or a supervised population level approach (i.e. identifying characteristics that are differentially abundant across case/control status). For the unsupervised approach, utilizing the unadjusted CBEA with the mixture normal distribution provides a good initial starting point. In the case where researchers only want to screen samples with mean-inflated taxon sets (instead of additionally detecting taxon sets with increased correlation), they can apply the adjusted approach, which can be effective at conserving type I error even for high correlation scenarios. However, the trade off for this adjustment is power, which decreases with increasing correlation. For the supervised analysis, all CBEA variants control for type I error and provide adequate statistical power. However, using raw CBEA scores with difference-in-means test such as Welch's t-test is preferable since is the least computationally expensive (no estimation process) while still outperforming the use of a sum-based approach with a standard differential abundance test.
